---
title: "STAT 2857A -- Lecture 16 Examples and Exercises"
subtitle: "Solutions -- Revised 7/11/24"
format: 
  pdf: 
   include-in-header: 
    - text: |
        \usepackage{mathtools}
knitr: 
  opts_chunk:
    fig.align: center
    echo: false
    warning: false
    message: false
    digits: 2
---

```{r}
## Load packages
library(tidyverse)
library(knitr)
library(latex2exp)
```

## Review

a) FALSE

b) FALSE

c) TRUE

d) This is a trick question. All of these are equal.

## Example 16.1 

a) 
$$
f(x)=\frac{1}{3\sqrt{2\pi}} \exp\left(-\frac{(x-64)^2}{18}\right)
$$
for all $x \in \Re$.

b) Note that 5~ft is equivalent to 60 inches and 6~ft to 72 inches. Plugging these values into the calculator on stattrek we find
$$
\begin{aligned}
      P(X < 60)=.09121
\end{aligned}
$$
and 
$$
P(X > 72) = .00383.
$$
Finally
$$
P(60 < X < 72)=.99612-.09121=.9049.
$$

c) Let $\eta_p$ denote the $(100p)-th$ percentile of the distribution, as we did before. Then 
$$
P(X \leq \eta_p)=p
$$
and
$$
P(X \leq \eta_{.95+p})=.95 + p
$$
for any $0 < p < .05$. Hence
$$
P(\eta_p < X < \eta_{.95+p})=(.95+p)-p=.95.
$$
The shortest interval will occur when the interval is symmetric about the mean since it will then contain the points with highest density. This occurs when $p=.025$. In that cas
$$
l=\eta_{.025}=58.120 \mbox{ and } u=\eta_{.0975}=69.880.
$$

d) 
$$
\begin{aligned}
      P(X \leq 60)
      &=P\left(\frac{X-64}{3} \leq \frac{60-64}{3}\right)\\
      &=P\left(Z \leq \frac{60-64}{3}\right)\\
      &=P(Z \leq -4/3)\\
      &=.09121.
\end{aligned}
$$

$$
\begin{aligned}
      P(X \geq 72)\\
      &=1-P(X \leq 72)\\
      &=1-P(\left(\frac{X-64}{3} \leq \frac{72-64}{3}\right)\\
      &=1-P\left(Z \leq \frac{72-64}{3}\right)\\
      &=1-P(Z \leq 8/3)\\
      &=1-.99617\\
      &=.00383
\end{aligned}
$$

e) Let $0<\alpha<.05$. Then 
$$
P(Z > z_\alpha)=\alpha
$$
and
$$
P(Z > z_{.95+\alpha})=P(Z>z_{.95+\alpha})=.95+\alpha.
$$
Then
$$
P(z_{.95+\alpha} < Z < z_\alpha)=(.95+\alpha)-\alpha=.95.
$$
Reversing the standardization, this means that
$$
P(\mu + (z_{.95+\alpha})\sigma < X < \mu + z_\alpha \sigma)=.95.
$$
Once again, the shortest interval will occur when the points are symmetric about the mean which happens when $\alpha=.025$. Applying the empirical rule\footnote{The exact values are -1.960 and 1.960.},
$$
z_{.975} \approx -2 \mbox{ and } z_{.025}\approx 2.
$$
Then
$$
l \approx 64 + 3(-2)=58 \mbox{ and } u \approx 64 + 3(2)=70.
$$

## Example 16.2

a) We know from the previous lecture that 
$$
X \sim \mbox{Binomial}(200,.486).
$$ 
Then
$$
\begin{aligned}
    P(95 \leq X \leq 105)
    &=\sum_{x=95}^{105} \binom{200}{y} .486^y.514^{200-y}\\
    &=`r dbinom(95,200,18/37)` +
      `r dbinom(96,200,18/37)` +
      \cdots +
      `r dbinom(104,200,18/37)` +
      `r dbinom(105,200,18/37)` \\
    &=`r sum(dbinom(95:105,200,18/37))`
\end{aligned}
$$
  
b) Alternatively, if we let $Z$ denote a standard normal random variable then 
$$
  \begin{aligned}
    P(95 \leq X \leq 105)
    &=P(X \leq 105) - P(X \leq 94)\\
    &\approx P\left(Z \leq \frac{105.5 - 200(.486)}{\sqrt{200(.486)(.514)}}\right) -
      P\left(Z \leq \frac{94.5 - 200(.486)}{\sqrt{200(.486)(.514)}}\right)\\
    &=P(Z \leq `r (u <- (105.5 - 200*(.486))/(sqrt(200*(.486)*(.514))))`) -
      P(Z \leq `r (l <- (94.5 - 200*(.486))/(sqrt(200*(.486)*(.514))))`)\\
    &=`r pnorm(u)` - `r pnorm(l)`\\
    &=`r pnorm(u)-pnorm(l)`
  \end{aligned}
$$
Although the probability is not exact it is very close.

## Exercise 16.1

a) 
```{r}
## Parameters
mu <- 28
sigma <- 4

## Create data frame
mydf <- tibble(x = mu + sigma * c(seq(-3,3,length=100)),
               f = dnorm(x, mu, sigma))

## Plot density
plot1 <- mydf %>%
  ggplot(aes(x = x, y = f)) +
  geom_line() +
  ylab("f(x)") +
  geom_vline(xintercept = mu, lty = 2)

print(plot1)
```

b) 
```{r}
## Plot density with probability area
xval <- 36
area1 <- filter(mydf, x >= xval) %>%
  bind_rows(tibble(x = c(max(mydf$x), xval, xval),
                   f = c(0,0, dnorm(xval,mu,sigma))))

plot1 +
  geom_polygon(data = area1, aes(x= x, y =f), alpha = .5)
```

c) Using the calculator
$$
P(X > 34)=.06681.
$$
Standardizing
$$
\begin{aligned}
P(X > 34)
&=1-P(X \leq 34)\\
&=1-P\left(Z \leq \frac{34 - 28}{4}\right)\\
&=1-P(Z \leq 1.5)\\
&=1-.9332\\
&=\textcolor{red}{.0668}.
\end{aligned}
$$

d) Here we can apply the empirical rule. Note that
$$
36=28 + 8=\mu + 2\sigma.
$$
Hence
$$
P(X > 36)=P(X >  \mu + 2\sigma)=(1-.95)/2=.025.
$$