---
title: "STAT 2857A -- Lecture 23 Examples and Exercises"
format: 
  pdf: 
   include-in-header: 
    - text: |
        \usepackage{mathtools}
knitr: 
  opts_chunk:
    fig.align: center
    echo: false
    warning: false
    message: false
    digits: 2
---
```{r}
## Load packages
library(tidyverse)
library(knitr)
library(latex2exp)
library(kableExtra)
library(readxl)

## Data summaries
apple_mu <- 0.1202
apple_var <- 5.2413
apple_sd <- 2.2984
apple_cov <- 0.0412
apple_cor <- apple_cov/apple_sd^2

## Options
opt_bkp <- options
options(digits = 4)
```

```{r}
mydata <- tibble(w = 0:3,
                 N = pmax(w,3-w)) |> 
  group_by(w,N) |>  
  reframe(r = 0:N) |> 
  mutate(`P(W=w)`=dhyper(w,18,9,3),
         `P(R=r|W=w)`=dhyper(r,26,26,N),
         `P(W=w,R=r)`=`P(W=w)`*`P(R=r|W=w)`) 
```

Let's play a game! 

Suppose that I have 

1) a bag that contains 18 white marbles and 9 black marbles, and 
2) a standard deck of cards. 

You will choose 3 marbles from the bag without replacement. Let $W$ be the number of white marbles. You will then draw $N=\max(W,3-W)$ cards from the deck without replacement. Let $R$ be the number of red cards. You win the game (and a prize) if all $N$ cards are the same colour -- i.e., if $R=N$ or $R=0$. 

The joint probability mass function of $W$ and $R$ has the following values:
```{r}
mydata |> 
  select(w,r,`P(W=w,R=r)`) |> 
  complete(w,r, fill = list(`P(W=w,R=r)` = 0)) |> 
  pivot_wider(names_from = r, values_from = `P(W=w,R=r)`) |>
  kable(digits = 3, align = "ccccc") |> 
  add_header_above(c("","r"=4), line = FALSE)
```

a) Show that the probability that $W=2$ and $R=1$ is .240?
b) Find a general expression for the joint probability mass function of $W$ and $R$?
c) What are the mean and variance of $R$?
d) What is the probability that you win?
e) What is the conditional pdf of $W$ given $R=2$?
f) What is the mean of $W$ given $R=2$? Provide an interpretation for this value.
g) What is the variance of $W$ given $R=2$?
h) What are the covariance and correlation of $W$ and $R$?