---
title: "STAT 2857A -- Lecture 23 Examples and Exercises"
subtitle: "Solutions -- Revised 27/11/24"
format: 
  pdf: 
   include-in-header: 
    - text: |
        \usepackage{mathtools}
knitr: 
  opts_chunk:
    fig.align: center
    echo: false
    warning: false
    message: false
    digits: 2
---
```{r}
## Load packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(latex2exp)
library(readxl)

## Options
opt_bkp <- options
options(digits = 3)
```

a) First consider the number of white marbles you draw from the bag. We know that $W$ will follow a hypergeometric distribution. Specifically
$$
W \sim \mbox{Hypergeometric}(3,18,27).
$$
Then
$$
P(W=w)=\frac{{18 \choose w}{9 \choose 3-w}}{{27 \choose 3}}.
$$
Hence,
$$
P(W=2)=\frac{{18 \choose w}{9 \choose 3-w}}{{27 \choose 3}}=`r (pw2 <- dhyper(2,18,9,3))`.
$$
Now consider the number of red cards that you draw from the deck. Given $W=2$, number of cards you choose is $N=\max(2,1)=2$. Then
$$
R|W=2 \sim \mbox{Hypergeometric}(2,26,52).
$$
So
$$
P(R=1|W=2)=\frac{{26 \choose 1}{26 \choose 1}}{{\textcolor{red}{52} \choose 2}}
=`r (pr1w2 <-dhyper(1,26,26,2))`.
$$
Finally,
$$
\begin{aligned}
P(W=2,R=1)
&=P(W=2)P(R=1|W=2)\\
&=`r pw2`(`r pr1w2`)\\
&=`r pw2 * pr1w2`.
\end{aligned}
$$

b) Again, we start with the number of white marbles you draw from the bag. As above,
$$
P(W=w)=\frac{{18 \choose w}{9 \choose 3-w}}{{27 \choose 3}}.
$$
Then, given $W=w$, $N=\max(w,3-w)$ and
$$
R|W=w \sim \mbox{Hypergeometric}(\max(w,3-w),26,52).
$$
So
$$
P(R=r|W=w)=\frac{{26 \choose r}{26 \choose \max(w,3-w)-r}}{{\textcolor{red}{52} \choose \max(w,3-w)}}.
$$
Finally, we consider the set of possible values (support) of $W$ and $R$. The possible values of $W$ are 0,1,2, and 3, and the possible values of $R$ are 0, 1, 2, and 3 if $W=0$ or $W=3$ and 0,1, or 2 if $W=1$ or $W=2$. Hence,
$$
P(W,R)=P(W=w)P(R=r|W=w)=\frac{{18 \choose w}{9 \choose 3-w}{26 \choose r}{52 \choose \max(w,3-w)-r}}{{27 \choose 3}{\textcolor{red}{52} \choose \max(w,3-w)}}.
$$
The following table provides the values in the intermediate calculations used to compute the table provided in the exercises:
```{r}
mydata <- tibble(w = 0:3,
                 N = pmax(w,3-w)) |> 
  group_by(w,N) |>  
  reframe(r = 0:N) |> 
  mutate(`P(W=w)`=dhyper(w,18,9,3),
         `P(R=r|W=w)`=dhyper(r,26,26,N),
         `P(W=w,R=r)`=`P(W=w)`*`P(R=r|W=w)`) 

mydata |> 
  kable(digits = 3)
```

b) To answer this question, we first need to find the marginal pmf for $R$. This is computed by summing the joint pdf over $w$ for each value of $r$. Equivalently, we sum the columns in the table above. The values are
```{r}
mydata1 <- mydata |>
  select(r,`P(W=w,R=r)`) |> 
  group_by(r) |> 
  summarise(`P(R=r)`=sum(`P(W=w,R=r)`)) 

erch <- paste0(mydata1$r,"(",round(pull(mydata1,2),3),")",collapse = "+")
er2ch <- paste0(mydata1$r^2,"(",round(pull(mydata1,2),3),")",collapse = "+")

er <- sum(pull(mydata1,1) * pull(mydata1,2))
er2 <- sum(pull(mydata1,1)^2 * pull(mydata1,2))
vr <- er2 - er^2

mydata1 |> 
  kable()
```
Then
$$
\begin{aligned}
E(R)&=`r erch`=`r er`,\\
E(R^2)&=`r er2ch`=`r er2`, \mbox{and}\\
V(R)&=E(R^2)-E(R)^2=`r vr`.
\end{aligned}
$$



c) You win if $R=0$ or $R=N$. Note that these are mutually exclusive. The probability that you win is
```{r}
pr0 <- mydata |> 
  filter(r==0) |> 
  pull(`P(W=w,R=r)`) 

pr0ch <- pr0 |>
  round(3) |> 
  paste(collapse = "+")

pr2 <- mydata |> 
  filter(r==2) |> 
  pull(`P(W=w,R=r)`) 

pr2ch <- pr2 |>
  round(3) |> 
  paste(collapse = "+")

pr3 <- mydata |> 
  filter(r==3) |> 
  pull(`P(W=w,R=r)`) 

pr3ch <- pr3 |>
  round(3) |> 
  paste(collapse = "+")

prN <- mydata |> 
  filter(r == w) |> 
  pull(`P(W=w,R=r)`) 

prNch <- prN |>
  round(3) |> 
  paste(collapse = "+")


```
$$
\begin{aligned}
P(R=0) + P(R=N)
=&(`r pr0ch`) + \textcolor{red}{(`r prNch`)}\\
=&\textcolor{red}{`r sum(pr0) + sum(prN)`}
\end{aligned}
$$

d) For any $w$ and $r$, the conditional probability of $W=w|R=r$ is
$$
P(W=w|R=r)=\frac{P(W=w,R=r)}{P(R=r)}.
$$
For the specific case of interest, we have
$$
P(R=2)=`r pr2ch` = `r pr2`.
$$
Then
$$
P(W=w|R=2)=\frac{P(W=w,R=2)}{`r sum(pr2)`}
$$
The entries in the conditional pdf are shown in the following table:
```{r}
mydata2 <- mydata |> 
  filter(r==2) |> 
  select(w,r,`P(W=w,R=r)`) |> 
  mutate(`P(W=w|R=2)`=`P(W=w,R=r)`/sum(pr2)) 

mydata2 |> 
  kable(digits = 3)
```

e) Using the values from the previous parts
```{r}
ewch <- paste0(mydata2$w,"(",round(pull(mydata2,4),3),")",collapse = "+")
ew2ch <- paste0(mydata2$w^2,"(",round(pull(mydata2,4),3),")",collapse = "+")

ew <- sum(pull(mydata2,1) * pull(mydata2,4))
ew2 <- sum(pull(mydata2,1)^2 * pull(mydata2,4))
```
$$
\begin{aligned}
E(W|R=2)&=`r ewch`=`r ew`
\end{aligned}
$$

f) Using the values from the previous parts
$$
\begin{aligned}
E(W^2|R=2)&=`r ew2ch`=`r ew2`
\end{aligned}
$$
and
$$
\begin{aligned}
V(W|R=2)&=E(W^2|R=r)-E(W|R=r)^2=`r ew2 - ew^2`.
\end{aligned}
$$

```{r}
ew <-  sum(mydata$w * pull(mydata,6))
vw <- (27-3)/(27-1)*3*2/3*1/3

erw <- sum(mydata$w * mydata$r *  pull(mydata,6))

mydata3 <- mydata |> 
  filter(w > 0, r>0)

erwch <- paste0(mydata3$w,"(",mydata3$r,")","(",round(pull(mydata3,6),3),")",collapse = "+")
```

g) The shortcut formula implies that
$$
\mbox{Cov}(W,R)=E(WR)-E(W)E(R).
$$
Since $W$ is hypergeometric, we know that 
$$
E(W)=\frac{3(18)}{27}=`r ew`.
$$
and
$$
V(W)=\frac{27-3}{27-1}\cdot\left(\frac{3(18)}{27}\right)\left(1-\frac{18}{27}\right)=`r vw`
$$
We also have that
$$
E(R)=`r er`.
$$
Then
$$
\begin{aligned}
E(RW)
&=`r erwch`\\
&=`r erw`.
\end{aligned}
$$
Hence
$$
\mbox{Cov}(W,R)
=`r erw` - (`r er`)(`r ew`)=`r (covrw <- erw - er*ew)`.
$$
Finally
$$
\begin{aligned}
\mbox{Corr}(W,R)
&=\frac{\mbox{Cov}(W,R)}{\sqrt{V(W)V(R)}}\\
&=\frac{`r covrw`}{\sqrt{(`r vw`)(`r vr`)}}\\
&=`r covrw/sqrt(vw * vr)`
\end{aligned}
$$