---
title: "STAT 2857A -- Lecture 17 Examples and Exercises"
subtitle: "Solutions"
format: 
  pdf: 
   include-in-header: 
    - text: |
        \usepackage{mathtools}
knitr: 
  opts_chunk:
    fig.align: center
    echo: false
    warning: false
    message: false
    digits: 2
---

```{r}
## Load packages
library(tidyverse)
library(knitr)
library(latex2exp)
```

## Review

a) FALSE (I've divided by the variance not the standard deviation.)
b) d
c) a
d) a

## Example 17.1

```{r}
alpha <- .628
beta <- 8.662
```

a) The pdf of $X$ is:
  $$
    f(x)=\frac{1}{`r beta`^{`r alpha`}\Gamma(`r alpha`)}x^{`r alpha`-1}e^{-x/`r beta`},\quad x \geq 0.
  $$

b) The mean and variance are:
$$
\begin{aligned}
    E(X)&=\alpha \beta = `r alpha`(`r beta`)=`r alpha * beta`\\
    V(X)&=\alpha \beta^2 =`r alpha`(`r beta`^2)=`r alpha * beta^2`\\
\end{aligned}
$$
  
c) The probability that the total precipitation is more than 10 mm given that it rains at all is:
$$
\begin{aligned}
    P(X \geq 10)
    &=1-P(X \leq 10)\\
    &=1-`r pgamma(10,alpha,1/beta)`\\
    &=`r 1-pgamma(10,alpha,1/beta)`
\end{aligned}
$$


## Example 17.2
```{r}
hlife <- .7038
lambda <- -log(.5)/hlife
p <- pexp(10^-9,lambda)
n <- 2.56e21

sci <- function(n) format(n, scientific = TRUE)
```

a) To write down the pdf we need first to understand what the half-life means. The half-life means that the probability of decay within the given time is .5 (i.e., the median). If we let $t_{1/2}$ denote the half-life then
$$
\begin{aligned}
    F(t_{1/2})&=.5\\
    1-e^{-\lambda t_{1/2}}&=.5\\
    \lambda t_{1/2}&=-\log(.5)\\
    \lambda&=`r -log(.5)/hlife`
  \end{aligned}
$$
The pdf is
$$
\begin{aligned}
    f(t)
    &=e^{-\lambda t}\\
    &=e^{-`r lambda` t}, \quad t > 0.
\end{aligned}
$$
  
b) The mean and variance of $T$ are
$$
  \begin{aligned}
    E(T)&=\frac{1}{\lambda}=\frac{1}{`r lambda`}=`r 1/lambda`\\
    V(T)&=\frac{1}{\lambda^2}=\frac{1}{`r lambda^2`}=`r 1/lambda^2`\\
  \end{aligned}
$$

c) The probability that $T>1$ is
$$
  \begin{aligned}
    P(T>1)
    &=1-P(T\leq 1)\\
    &=1-(1-e^{-`r lambda`})\\
    &=e^{-`r lambda`}\\
    &=`r 1-pexp(1,lambda)`
  \end{aligned}
$$
  
d) The probability that $T > 2$ given $T>1$ is
$$
  \begin{aligned}
    P(T>2 | T>1)
    &=\frac{P(T>2,T>1)}{P(T>1)}\\
    &=\frac{P(T>2)}{P(T>1)}\\
    &=\frac{e^{-2(`r lambda`)}}{e^{-(`r lambda`)}}\\
    &={e^{-(`r lambda`)}}\\
    &=`r 1-pexp(1,lambda)`\\
    &=P(T>1)
  \end{aligned}
$$
e) The probability that $T > 100,001$ given $T>100,000$ is
$$
  \begin{aligned}
    P(T>100,001 | T>100,000)
    &=\frac{P(T>100,001)}{P(T>100,000)}\\
    &=\frac{e^{-100,001(`r lambda`)}}{e^{-100,000(`r lambda`)}}\\
    &={e^{-(`r lambda`)}}\\
    &=`r 1-pexp(1,lambda)`\\
    &=P(T>1)
  \end{aligned}
$$
     In fact, for any $t_0,t >0$
$$
    P(T>t+t_0|T>t_0)=P(T>t).
$$

## Example 17.3

Suppose that $Z \sim N(0,1)$ then
$$
\begin{aligned}
  P(Z^2 \leq 2)
  &=P(-sqrt{2} < Z < \sqrt{2})\\
  &=P(-1.4142 < Z < 1.4142)\\
  &=P(Z < 1.4142) - P(Z < -1.4142)\\
  &=.9213 - .0787\\
  &=.8426.
\end{aligned}
$$

Using the online calculator at stattrek we find that
$$
  P(X < 2) = .8426
$$
exactly as required. In fact, this would hold for any value, not just 2. 

## Exercise 17.1

a) According to the question,
$$
T_5 \sim \mbox{Gamma}(5,10).
$$
Then $\mu_{T_5}=E(T_5)=50$ and $\sigma^2_{T_5}=V(T_5)=500$. 

b) Using the online calculator
$$
P(T_5 > 60)=0.28506.
$$

c) Using the calculator, the 5-th, 50-th (median), and 95-th percentiles of $T_5$ and the normal distribution with mean $50$ and variance $500$ are
```{r}
mydat <- tibble(p = c(5, 50, 95)/100,
                Gamma = qgamma(p, 5, scale = 10),
                Normal = qnorm(p, 50, sqrt(500)))
```

d) The following plot compares the ratio of the 5-th, 50-th, and 95-th percentiles for the gamma distribution with $\alpha = k$ and $\beta = 10$ and the normal distribution with matching mean and variance ($\mu=10k$ and $\sigma^2=100k$). For all three percentiles, the ratio gets closer to 1.00 as $k$ increases. This implies that the percentiles of the gamma and the percentiles of the normal distribution are becoming closer and closer, relatively speaking.
```{r}
mydat <- crossing(p = c(5, 50, 95)/100,
                  k = c(5, 25, 50, 100)) |> 
  mutate(Gamma = qgamma(p, k, scale = 10),
         Normal = qnorm(p, k*10, 10*sqrt(k)),
         r = Gamma/Normal)

mydat |> 
  ggplot(aes(x = k, y = r, colour = as.factor(100*p))) + 
  geom_point() + 
  ylim(c(0,NA)) + 
  geom_hline(yintercept = 1, lty = 2) +
  ylab("Ratio of Percentiles (Gamma/Normal)") +
  labs(colour = "Percentile")
```
  