---
title: "STAT 2857A -- Lecture 14 Examples and Exercises"
subtitle: "Solutions -- Revised 31/10/24"
format: 
  pdf: 
   include-in-header: 
    - text: |
        \usepackage{mathtools}
knitr: 
  opts_chunk:
    fig.align: center
    echo: false
    warning: false
    message: false
    digits: 2
---

```{r}
## Load packages
library(tidyverse)
library(knitr)
library(latex2exp)
```

## Example 14.1

a) To show that $f(x)$ is a proper density function we need to show that:

  i. $f(x) \geq 0$ for all $x \in \Re$
  ii. $\int_{\infty}^\infty f(u)~du=1$

  The first criterion requires that $c \geq 0$. Then we need
  $$
  \begin{aligned}
    \int_{\infty}^\infty f(u)~du &= 1\\
    \int_{-\infty}^0 0~du + \int_0^1 cu^2~du + \int_1^\infty 0~du & = 1\\
    \left[\frac{cu^2}{2}\right]_0^1& =1\\
      c&=2
    \end{aligned}
  $$
  Hence, the pdf is proper if $c=2$. This give us
  $$
  f(x)=\left\{
    \begin{array}{ll}
      0 \leq 0\\
      2x & 0 < x < 1\\
      0 & 1 \leq x
    \end{array}
  \right.
  $$
  
b) By definition, the cdf is
  $$
  F(x)=\int_{-\infty}^x f(u)~du=
    \left\{
        \begin{array}{ll}
          0 & x \leq 0\\
          x^2 & 0 < x < 1\\
          1 & 1 \leq x
        \end{array}
      \right.
  $$
  
c) 

  i) By definition:
  $$
      \begin{aligned}
        P(X \leq .5) 
        &=F(.5)\\
        &=.5^2\\
        &=.25
      \end{aligned}
  $$

  ii) For any continuous random variable $P(X=x)=0$ for any $x \in \Re$. Hence, $P(X=.5)=0$. 

  iii) Consider that:
    $$
    \begin{aligned}
        P(X < .5)
        &=P(X \leq .5) - P(X=.5)\\
        &=F(.5) - 0\\
        &=.5^2\\
        &=.25
    \end{aligned}
    $$
   Note that $P(X < .5)=P(X \leq .5)$. In general, if $X$ is a continuous random variable then $P(X<x)=P(X\leq x)=F(x)$. 
    
  
  iv) 
  $$
      \begin{aligned}
        P(.25 \leq X \leq .75)
        &=P(X \leq .75) - P(X < .25)\\
        &=P(X \leq .75) - P(X \leq .25)\\
        &=.75^2-.25^2\\
        &=.50
      \end{aligned}
  $$

  v)
  $$
      \begin{aligned}
        P((X <.25) \cup (X >.75)
        &=1-P(.25 \leq X \leq .75)\\
        &=1-[F(.75)-F(.25)]\\
        &=1-F(.75) + F(.25)\\
        &=1-.75^2+.25^2\\
        &=1-.5625+.0625
      \end{aligned}
  $$
  
d) The random variable $X$ is continuous, according to Devore and Berk, if:
    
   i) the possible values (support) of $X$ is a union of (possibly one) disjoint interval in $\Re$, and
   ii) $P(X=x)=0$ for any $x \in \Re$. 

  The first criterion is  satisfied because the support of $X$ is a single interval, $(0,1)$.

  To show that the second criterion is true consider that
    $$
    \begin{aligned}
      P(X=x)
      &=P(X \leq x) - P(X < x)\\
      &=P(X \leq x) - \lim_{x^- \nearrow} P(X \leq x^{-})\\
      &=F(x) - \lim_{x^- \nearrow x} F(x^-)\\
      &=x^2 - \lim_{x^- \nearrow x} (x^-)^2\\
      &=0.
    \end{aligned}
    $$
  What the second criterion really implies is that $F(x)$ is continuous. Hence a continuous random variable!
  
## Exercise 14.2

We know from Example 14.1. that the cdf is
$$
  F(x)=\int_{-\infty}^x f(u)~du=
    \left\{
        \begin{array}{ll}
          0 & x \leq 0\\
          x^2 & 0 < x < 1\\
          1 & 1 \leq x
        \end{array}
      \right.
$$
Inverting the cdf we have that
$$
\eta_p=\sqrt{p}
$$
for any $0 \leq p \leq 1$ where we are considering the positive square root. 

a) The median is
$$
\eta_{.5}=\sqrt{.5}=.7071.
$$

b) The 5th and 95th percentiles are
$$
\begin{aligned}
\eta_{.05}&=\sqrt{.05}=.2236\\
\eta_{.95}&=\sqrt{.95}=.9747.
\end{aligned}
$$

c) Since $f(x)$ is monotonically increasing where it is positive, $(0,1)$, the shortest interval will occur when $x_2=1$. Then $x_1$ must satisfy
$$
P(x_1 < X <1)=.90
$$
which implies that
$$
P(X < x_1)=.10
$$
or 
$$
x_1=\eta_{.10}=\sqrt{.10}=.3162.
$$

## Exercise 14.1

a)

```{r}
mydat1 <- tibble(x = 9*(0:100)/100) |> 
  mutate(f = 1/((x + 1) * log(10)),
         F = log10(x + 1)) |> 
  arrange(x)

mydat <- tibble(x = c(-1,0), f = c(0,0), F = c(0,0)) |> 
  bind_rows(mydat1) |> 
  bind_rows(tibble(x = c(9,10), f =c(0,0), F= c(1,1)))

mydat |> 
  ggplot(aes(x = x, y = F)) +
  geom_line() +
  xlab("x") + 
  ylab("F(x)")
```

b) The pdf is given by
$$
\begin{aligned}
f(x)
&=\frac{d}{dx} F(x).\\
\end{aligned}
$$
Note that $F(x)$ is flat over the regions $x<0$ and $x>9$ so
$$
f(x) = \frac{d}{dx} F(x)=0.
$$
For $0 \leq x \leq 9$
$$
\begin{aligned}
f(x)
&=\frac{d}{dx} \log_10(1 + x)\\
&=\frac{1}{(x + 1)\ln(10)}. 
\end{aligned}
$$
c)

```{r}
mydat |> 
  ggplot(aes(x = x, y = f)) +
  geom_line() +
  xlab("x") + 
  ylab("f(x)")
```

d) 

    i) $P(X \leq \sqrt{10}\textcolor{red}{-1}) = F(\sqrt{10}\textcolor{red}{-1}) = \log_{10}(\sqrt{10})=.50$.
    ii) Since the cdf is continuous, $P(X < \sqrt{10}\textcolor{red}{-1})=P(X \leq \sqrt{10}\textcolor{red}{-1})=.50$. 
    iii) Since the cdf is continuous, $P(X=\sqrt{10}\textcolor{red}{-1})=0$.
    iv) $P(X > \sqrt{10}\textcolor{red}{-1})=1-P(X \leq \sqrt{10}\textcolor{red}{-1})=.5$
  v)$P(X \geq \sqrt{10}\textcolor{red}{-1})=1-P(X < \sqrt{10}\textcolor{red}{-1})=.5$
  